---
title: "Multiple linear regression - Cholesterol"
format: html
editor: visual
---

## =============================================================================

## MHEDAS

## Biomedical Statistics

## Multiple linear regression

## Outcome: cholesterol (C)

## Explanatory variables: height in cm (H); Weight in Kg (W); Age in years (A)

## Main concepts: confounders and Collinearity

## =============================================================================

## Load libraries

## -----------------------------------------------------------------------------

```{r}
library(car) # Linear modeling
library(HH)  # Data display
```

## Read data

## -----------------------------------------------------------------------------

```{r}
dd <- read.csv2("cholesterol.csv") # dataset
n  <- dim(dd)[1]                   # sample size
head(dd)
```

## Descriptive plot

## -----------------------------------------------------------------------------

HH package sp()

```{r}
sp(C ~ W, boxplot=F, smooth=F, dd)
```

relationship looks wrong (weight, cholesterol) -\> check for confounders

## Null model

## -----------------------------------------------------------------------------

```{r}
mod0 <- lm(C~1,dd) #simple model
summary(mod0)
```

```{r}

## Simple linear regression with weight 
## -----------------------------------------------------------------------------
mod <- lm(C ~ W, dd)# Fit the model with weight, simple model
summary(mod)        # Summary of the model
anova(mod0,mod)     # Test Omnibus to check if the model explains something on the outcome
```

cholesterol goes down with higher weight

## Predictions

## -----------------------------------------------------------------------------

```{r}
CC <- predict(mod, interval="confidence", alpha=0.01)  # Expected value, avarege of all
CP <- predict(mod, interval="prediction", alpha=0.01)  # Point prediction, one individual observation
```

## -- Plot

```{r}
O <- order(dd$W)                                     # Order
plot(dd$W,dd$C,ylim=c(min(CP,dd$C),max(CP,dd$C)),
     ylab="C",xlab="P",main="99% prediction bands")  # Points

lines(dd$W[O],CC[O,"fit"],lty=2,col="blue")          # Point estimate

lines(dd$W[O],CC[O,"lwr"],lty=2,col="blue")          # Lower limit confidence interval
lines(dd$W[O],CC[O,"upr"],lty=2,col="blue")          # Upper limit confidence interval

lines(dd$W[O],CP[O,"lwr"],lty=2,col="red")           # Lower limit prediction interval
lines(dd$W[O],CP[O,"upr"],lty=2,col="red")           # Upper limit prediction interval
```

## -- Alternative plot

```{r}
ci.plot(mod,conf.level=.99) 
```

## Validation

## -----------------------------------------------------------------------------

##-- Linearity and homoscedasticity

```{r}
plot(predict(mod),resid(mod),main="Residuals vs Predicted values")
abline(h=0,lty=2)
```

##-- Normality of the residuals

```{r}
qqnorm(resid(mod))
qqline(resid(mod),col=2)
```

##-- Outliers

```{r}
plot(rstudent(mod),main="Studentized Residuals")
abline(h=c(-3,-2,0,2,3),lty=2)
```

one observation is very high, far away

##-- Influential observations (Cook's distance)

```{r}
plot(mod,4)
abline(h = 4/nrow(dd), lty=2, col='grey')
```

one observation has a very high influence

##-- Influential observations (dffits is an alternative measure of the influence)

```{r}
p <- 2
plot(dffits(mod),main="dffits")
abline(h=c(-3,-2,0,2,3)*sqrt(p/n),lty=2)
```

##-- All in one

```{r}
par(mfrow=c(2,2))
plot(mod)
```

## Problem with the interpretation

checking counfounders \## -----------------------------------------------------------------------------

```{r}
scatterplot(C~W|A,smooth=F,boxplots=F,col=1:12,dd)
```

-   one symbol = one age
-   almost all regression lines = positive slopes
-   ages are causing overall neg slope

## Multiple linear regression with all variables

## -----------------------------------------------------------------------------

##-- Fit the model

```{r}
mod3v <- lm(C ~ W + A + H,dd)
summary(mod3v)
```

coefficient of W = positive

##-- Test significance

```{r}
anova(mod3v)   # Sequential relevance (Anova Type I)
```

order as order of variables in the model

```{r}
Anova(mod3v)   # Marginal relevance   (Anova type II) 
```

\`\*´= relevant variable

##-- Collinearity

```{r}
vif(mod3v)
```

## Validation

## -----------------------------------------------------------------------------

##-- Linearity and homoscedasticity

```{r}
plot(predict(mod3v),resid(mod3v),main="Residuals vs Predicted values")
abline(h=0,lty=2)
```

##-- Normality of the residuals

```{r}
qqnorm(resid(mod3v))
qqline(resid(mod3v),col=2)
```

##-- Outliers

```{r}
plot(rstudent(mod3v),main="Studentized Residuals")
abline(h=c(-3,-2,0,2,3),lty=2)
```

##-- Influential observations (Cook's distance)

```{r}
plot(mod3v,4)
abline(h = 4/nrow(dd), lty=2, col='grey')
```

##-- All in one

```{r}
par(mfrow=c(2,2))
plot(mod3v)
```

## Dealing with collinearity --\> Create variable EW: excess weight

defined by expert: \## Ideal weight = 0.5 \* height - 10 \## ----------------------------------------------------------------------------- \# New variable include info of weight and height

```{r}
real_weight  <- dd$W
ideal_weight <- dd$H/2 - 10
dd$EW <- pmax(0, real_weight - ideal_weight)
```

## Multiple linear regression with excess weight

## -----------------------------------------------------------------------------

```{r}
modE2v <- lm(C ~ EW + A, dd)
summary(modE2v)
```

## Comparing predictions

## -----------------------------------------------------------------------------

##-- Comparing (absolute/relative) predictions

```{r}
par(mfrow=c(1,2))
plot(predict(mod3v) - predict(modE2v),
     main="Absolute differences \n in predictions")
abline(h=0,lt=2)
plot((predict(mod3v) - predict(modE2v))/predict(mod3v),
     main="Relative differences \n in predictions")
abline(h=0,lt=2)
```

##-- Comparing residual standard deviation

```{r}
(res_sd <- c(summary(mod3v)$sigma,
             summary(modE2v)$sigma))
```

##-- Comparing coefficients

```{r}
rbind(coef(mod3v),c(coef(modE2v),NA))
```

NA = independence; solved problem of collinearity

##-- Tests significance

```{r}
anova(modE2v)   # Sequential contribution
```

```{r}
Anova(modE2v)   # Marginal contribution
```

##-- Collinearity

```{r}
vif(modE2v)          
```

## Validation

## -----------------------------------------------------------------------------

##-- All in one

```{r}
par(mfrow=c(2,2))
plot(modE2v)
```

```{r}
residualPlots(modE2v)
```

## Predictions

## -----------------------------------------------------------------------------

##-- Prediction according to age for patients without excess of weight

```{r}
age  <- 9:20
C1   <- data.frame(cbind(A=age,EW=0))              # Individual characteristics
CC1  <- predict(modE2v, C1, interval="confidence") # 95% Confidence intervals
CP1  <- predict(modE2v, C1, interval="prediction") # 95% Prediction intervals

##-- Plot
par(mfrow=c(1,1))
plot (age,CC1[,"fit"],ty="l",ylim=c(min(CP1),max(CP1)),
      main="95% Prediction band for cholesterol 95% \n w/o excess of weight")

lines(age,CC1[,"lwr"],lty=2,col="blue")
lines(age,CC1[,"upr"],lty=2,col="blue")

lines(age,CP1[,"lwr"],lty=2,col="red")
lines(age,CP1[,"upr"],lty=2,col="red")
```

blue = average with 95% confidence for all people in the world

##-- Prediction for a patient with W=65, A=15 i H=150

```{r}
C0 <- data.frame(cbind(EW=c(max(0,65-(150/2-10))), A=c(15)))
predict(modE2v, C0, interval="confidence", level=.95) # confidence interval
predict(modE2v, C0, interval="prediction", level=.95) # prediction interval
```

## Contrasts

## -----------------------------------------------------------------------------

##-- Contrast if E\[C\]=190 for the previous individual with W=65, A=15 i H=150 optipon. 0 = see if value could be 190 -\> no cause it is is outside CI \# 1st option: manual

```{r}
pC <- predict(modE2v,C0,se.fit=T)
2*(1 - pt((pC$fit-190)/pC$se.fit, pC$df)) # p value
```

\<0.05 = significant difference to predicted average (200.043)

# 2nd option: "linearHypothesis" function

```{r}
linearHypothesis(modE2v,c(1,0,15),190)
```

# 3rd option: "lht" function

```{r}
lht(modE2v,c(1,0,15),190)
((pC$fit-190)/pC$se.fit)^2
```
